```{r message=FALSE, warning=FALSE}
library(imager)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(momentuHMM)
library(expm)
library(MASS)
library(clock)
library(qdapRegex)
library(mixR)
library(kableExtra)
library(pbs)
```



```{r}
#estimate initial parameters for all 3 states
#init_params30 <- initz(na.omit(df30$Depth), ncomp = 3, init.method = "hclust")
#init_params17 <- initz(na.omit(df17$Depth), ncomp = 3, init.method = "hclust")
#init_params6 <- initz(na.omit(df6$Depth), ncomp = 3, init.method = "hclust")
```



```{r}
# Initial values
### Depth
mean30_Depth <- c(220.7267738, 591.5, 831.0199)
sd30_Depth <- c(34.10027, 90.82657, 52.55956)
shape30_Depth <- c(15.94281, 87.588452)
scale30_Depth <- c(13.84491, 6.806144)
df30$Depth[1] <- 1

mean17_Depth <- c(79.70845, 281.24234, 421.93630)
sd17_Depth <- c(64.42877, 51.46055, 36.73780)
shape17_Depth <- c(13.6017578, 140.0003906)
scale17_Depth <- c(25.07972, 2.098351)

mean6_Depth <- c(321.6364, 692.8450, 1078.7261)
sd6_Depth <- c(76.50733, 72.92805, 146.37132)
shape6_Depth <- c(11.3389648, 40.0031250)
scale6_Depth <- c(28.83206, 18.39503)
```

# Null 2 states

```{r}
# distributions for observation processes
dist <- list(Depth = "gamma")
```

```{r}
### Shark 30

dat30 <- prepData(df30, coordNames = NULL)

# initial parameters
Par0_m30 <- list(Depth=c(mean30_Depth[1],mean30_Depth[2],
                               sd30_Depth[1],sd30_Depth[2]))
# fit model
m30 <- fitHMM(data = dat30, nbStates = 2, dist = dist, Par0 = Par0_m30)

```


```{r}
### Shark 17
dat17 <- prepData(df17, coordNames = NULL)

# initial parameters
Par0_m17 <- list(Depth=c(mean17_Depth[1],mean17_Depth[2],
                               sd17_Depth[1],sd17_Depth[2]))
# fit model
m17 <- fitHMM(data = dat17, nbStates = 2, dist = dist, Par0 = Par0_m17)

```

```{r}
### Shark 6
dat6 <- prepData(df6, coordNames = NULL)

# initial parameters
Par0_m6 <- list(Depth=c(mean6_Depth[1],mean6_Depth[2],
                              sd6_Depth[1],sd6_Depth[2]))
# fit model
m6 <- fitHMM(data = dat6, nbStates = 2, dist = dist, Par0 = Par0_m6)

```

# Null 3 states

```{r}
### Shark 30
# initial parameters
Par0_m30_3 <- list(Depth=c(mean30_Depth[1],mean30_Depth[2],mean30_Depth[3],
                               sd30_Depth[1],sd30_Depth[2],sd30_Depth[3]))
# fit model
m30_3 <- fitHMM(data = dat30, nbStates = 3, dist = dist, Par0 = Par0_m30_3)
```

```{r}
### Shark 17
dat17 <- prepData(df17, coordNames = NULL)

# initial parameters
Par0_m17_3 <- list(Depth=c(mean17_Depth[1],mean17_Depth[2],mean17_Depth[3],
                               sd17_Depth[1],sd17_Depth[2],sd17_Depth[3]))
# fit model
m17_3 <- fitHMM(data = dat17, nbStates = 3, dist = dist, Par0 = Par0_m17_3)

```

```{r}
### Shark 6
dat6 <- prepData(df6, coordNames = NULL)

# initial parameters
Par0_m6_3 <- list(Depth=c(mean6_Depth[1],mean6_Depth[2],mean6_Depth[3],
                               sd6_Depth[1],sd6_Depth[2],sd6_Depth[3]))
# fit model
m6_3 <- fitHMM(data = dat6, nbStates = 3, dist = dist, Par0 = Par0_m6_3)

```






# Covariate - Temperature - for transition probabilities - 2 states

```{r}
### Shark 30
#remove NA's
dat30$Temperature <- imputeTS::na_locf(dat30$Temperature)
# initial parameters (obtained from nested model m1)
Par0_m30_Temp <- getPar0(model=m30, formula=~Temperature)

# fit model with formula for transition probabilities
m30_Temp <- fitHMM(data = dat30, nbStates = 2, dist = dist, Par0 = Par0_m30_Temp$Par,
                        beta0=Par0_m30_Temp$beta, formula=~Temperature)

```



```{r}
### Shark 6
# remove NA
dat6$Temperature <- imputeTS::na_locf(dat6$Temperature)

# initial parameters (obtained from nested model m1)
Par0_m6_Temp <- getPar0(model=m6, formula=~Temperature)

# fit model with formula for transition probabilities
m6_Temp <- fitHMM(data = dat6, nbStates = 2, dist = dist, Par0 = Par0_m6_Temp$Par,
                        beta0=Par0_m6_Temp$beta, formula=~Temperature)
```



# Covariate - Temperature - for transition probabilities - 3 states


```{r}
### Shark 30
# remove NA
dat30$Temperature <- imputeTS::na_locf(dat30$Temperature)

# initial parameters (obtained from nested model m1)
Par0_m30_Temp_3 <- getPar0(model=m30_3, formula=~Temperature)

# fit model with formula for transition probabilities
m30_3_Temp <- fitHMM(data = dat30, nbStates = 3, dist = dist, Par0 = Par0_m30_Temp_3$Par,
                        beta0=Par0_m30_Temp_3$beta, formula=~Temperature)

```


```{r}
### Shark 6
# remove NA
dat6$Temperature <- imputeTS::na_locf(dat6$Temperature)

# initial parameters (obtained from nested model m1)
Par0_m6_Temp_3 <- getPar0(model=m6_3, formula=~Temperature)

# fit model with formula for transition probabilities
m6_3_Temp <- fitHMM(data = dat6, nbStates = 3, dist = dist, Par0 = Par0_m6_Temp_3$Par,
                        beta0=Par0_m6_Temp_3$beta, formula=~Temperature)
```


# Covariate - periodic spline of Time Of Day - for transition probabilities - 2 states 

```{r}
### Shark 30
dat30 <- prepData(df30, coordNames = NULL)
levels(dat30$ID) <- "shark 30"
# initial parameters (obtained from nested model m1)
Par0_m30_TOD <- getPar0(model=m30, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

# fit model
m30_TOD <- fitHMM(data = dat30, nbStates = 2, dist = dist, Par0 = Par0_m30_TOD$Par,
                        beta0=Par0_m30_TOD$beta, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))
```


```{r}
### Shark 17
# initial parameters (obtained from nested model m1)
Par0_m17_TOD <- getPar0(model=m17, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

# fit model
m17_TOD <- fitHMM(data = dat17, nbStates = 2, dist = dist, Par0 = Par0_m17_TOD$Par,
                        beta0=Par0_m17_TOD$beta, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

```

```{r}
### Shark 6 TOD 2 state
#getpar0 from m6 null model // intercept beta_0
Par0_m6_TOD <- getPar0(model=m6, 
                       formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

#Fithmm 
m6_TOD <- fitHMM(data = dat6, nbStates = 2, dist = dist,
                 Par0 = Par0_m6_TOD$Par,
                        beta0=Par0_m6_TOD$beta, 
                 formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))
```



# Covariate - periodic spline of Time Of Day - for transition probabilities - 3 states 

```{r}
### Shark 30
# initial parameters (obtained from nested model m1)
Par0_m30_3_TOD <- getPar0(model=m30_3, formula=~(splines2::mSpline(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

# fit model
m30_3_TOD <- fitHMM(data = dat30, nbStates = 3, dist = dist, Par0 = Par0_m30_3_TOD$Par,
                        beta0=Par0_m30_3_TOD$beta, formula=~(splines2::mSpline(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

```






```{r}
### Shark 17
# initial parameters (obtained from nested model m1)
Par0_m17_3_TOD <- getPar0(model=m17_3, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

# fit model
m17_3_TOD <- fitHMM(data = dat17, nbStates = 3, dist = dist, Par0 = Par0_m17_3_TOD$Par,
                        beta0=Par0_m17_3_TOD$beta, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))
```


```{r}
### Shark 6
# initial parameters (obtained from nested model m1)
Par0_m6_3_TOD <- getPar0(model=m6_3, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))

# fit model
m6_3_TOD <- fitHMM(data = dat6, nbStates = 3, dist = dist, Par0 = Par0_m6_3_TOD$Par,
                        beta0=Par0_m6_3_TOD$beta, formula=~(pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24))))
```



# Covariate - periodic time of year - for transition probabilities - 2 states 

```{r}
### Shark 30
dat30 <- prepData(df30, coordNames = NULL)
levels(dat30$ID) <- "shark 30"
# initial parameters (obtained from nested model m1)
Par0_m30_month <- getPar0(model=m30, formula=~(pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13))))

# fit model with formula for transition probabilities
m30_month <- fitHMM(data = dat30, nbStates = 2, dist = dist, Par0 = Par0_m30_month$Par,
                        beta0=Par0_m30_month$beta, formula=~(pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13))))
```



```{r}
### Shark 17
#initial parameters (obtained from nested model m1)
Par0_m17_month <- getPar0(model=m17, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))

# fit model with formula for transition probabilities
m17_month <- fitHMM(data = dat17, nbStates = 2, dist = dist, Par0 = Par0_m17_month$Par,
                        beta0=Par0_m17_month$beta, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))
```


```{r}
### Shark 6
# initial parameters (obtained from nested model m1)
Par0_m6_month <- getPar0(model=m6, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))

#fit model with formula for transition probabilities
m6_month <- fitHMM(data = dat6, nbStates = 2, dist = dist, Par0 = Par0_m6_month$Par,
                        beta0=Par0_m6_month$beta, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))
```


# Covariate - periodic time of year - for transition probabilities - 3 states 


```{r}
### Shark 30
#initial parameters (obtained from nested model m1)
Par0_m30_3_month <- getPar0(model=m30_3, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))

# fit model with formula for transition probabilities
m30_3_month <- fitHMM(data = dat30, nbStates = 3, dist = dist, Par0 = Par0_m30_3_month$Par,
                       beta0=Par0_m30_3_month$beta, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))
```



```{r}
### Shark 17
# initial parameters (obtained from nested model m1)
Par0_m17_3_month <- getPar0(model=m17_3, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))

# fit model with formula for transition probabilities
m17_3_month <- fitHMM(data = dat17, nbStates = 3, dist = dist, Par0 = Par0_m17_3_month$Par,
                      beta0=Par0_m17_3_month$beta, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))
```



```{r}
### Shark 6
# initial parameters (obtained from nested model m1)
Par0_m6_3_month <- getPar0(model=m6_3, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))

# fit model with formula for transition probabilities
m6_3_month <- fitHMM(data = dat6, nbStates = 3, dist = dist, Par0 = Par0_m6_3_month$Par,
                        beta0=Par0_m6_3_month$beta, formula=~pbs::pbs(TOY, df=3, periodic=TRUE, Boundary.knots=c(1,13)))
```


# temperature + TOD - 2 states

```{r}
### Shark 30
# initial parameters (obtained from nested model m1)
Par0_m30_TODTemp<- getPar0(model=m30, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))

# fit model
m30_TODTemp <- fitHMM(data = dat30, nbStates = 2, dist = dist, Par0 = Par0_m30_TODTemp$Par,
                        beta0=Par0_m30_TODTemp$beta, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))
```



```{r}
### Shark 6
# initial parameters (obtained from nested model m1)
Par0_m6_TODTemp<- getPar0(model=m6, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))

# fit model
m6_TODTemp <- fitHMM(data = dat6, nbStates = 2, dist = dist, Par0 = Par0_m6_TODTemp$Par,
                        beta0=Par0_m6_TODTemp$beta, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))

```


# temperature + TOD - 3 states
```{r}
### Shark 30
# initial parameters (obtained from nested model m1)
Par0_m30_3_TODTemp <- getPar0(model=m30_3, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))

# fit model
m30_3_TODTemp <- fitHMM(data = dat30, nbStates = 3, dist = dist, Par0 = Par0_m30_3_TODTemp$Par,
                        beta0=Par0_m30_3_TODTemp$beta, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))
```



```{r}
### Shark 6
# initial parameters (obtained from nested model m1)
Par0_m6_3_TODTemp <- getPar0(model=m6_3, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))

# fit model
m6_3_TODTemp <- fitHMM(data = dat6, nbStates = 3, dist = dist, Par0 = Par0_m6_3_TODTemp$Par,
                        beta0=Par0_m6_3_TODTemp$beta, formula=~((pbs::pbs(TOD, df=3, periodic=TRUE, Boundary.knots=c(0,24)))+Temperature))
```






